---
title: "Multiple Regression I"
subtitle: "Variable Selection"
author: "Cristian Mesquida"
institute: "Eindhoven University of Technology"
date: today
format: revealjs
progress: true
bibliography: references.bib
nocite: |
 @*
editor: visual
---

```{r packages}
#| include: false

library(dplyr)
library(ggplot2)
library(broom)
library(ggdag)
library(flextable)
```

## Abstract

This lecture introduces **variable selection in multiple regression**, with a focus on choosing predictors based on the research question. It first introduces **Directed Acyclic Graphs (DAGs)** and shows how they can help identify potential **confounders** and avoid inappropriate adjustment for **mediators** and **colliders**. Finally, it distinguishes **confirmatory research from exploratory research and prediction** and discusses how variable selection differs depending on the goal of the analysis.

```{r setup}
#| include: false

set_flextable_defaults(
  font.size = 20
)
```

## Simple regression

$$Y_i = \beta_0 + \beta_1X_{1i} + \epsilon_i $$

Where $Y_i$ = observed outcome, $\beta_0$ = intercept, $\beta_1$ = regression coefficient, $X_{1i}$ = predictor and $\epsilon_i$ = error term for observation $i$

**An example:**

$$Grade_i = \beta_0 + \beta_1 \times Intervention + \epsilon_i $$

Where **Intervention** is categorical (e.g., 0 = no intervention and 1 = intervention)

## Multiple regression

$$Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + ... +  \epsilon_i$$

Where $\beta_1, \beta_2, \dots$ = regression coefficients and $X_{1i}, X_{2i}, \dots$ = predictors for observation $i$

**An example:**

$$Grade_i = \beta_0 + 
\beta_1  \text{Intervention} + 
\beta_2  \text{Sleep_time} + 
\beta_3 \text{Study_time} + 
\epsilon $$

Where **Intervention** is categorical (e.g., 0 = no intervention and 1 = intervention) and **Sleep_time** and **Study_time** are continuous variables

## What variables should we include in the statistical model?

::: {.fragment .fade-in-then-out}
**Not** based on whether $p < 0.05$!

Selecting variables based solely on statistical significance is **a mindless approach** and should be avoided.
:::

::: {.fragment .fade-up}
It depends on the the **causal relationships** between variables and the **research goal**
:::

## Relationships between variables

```{r}
smoking_ca_dag <- ggdag::dagify(cardiacarrest ~ cholesterol,
                                cholesterol ~ smoking + weight,
                                smoking ~ unhealthy,
                                weight ~ unhealthy,
                                labels = c("cardiacarrest" = "Cardiac\n Arrest", 
                                           "smoking" = "Smoking",
                                           "cholesterol" = "Cholesterol",
                                           "unhealthy" = "Unhealthy\n Lifestyle",
                                           "weight" = "Weight")
                                )

ggdag(smoking_ca_dag, 
      text = FALSE,  
      use_labels = "label") + 
  theme_void()
```

## Direct Acyclic Graph (DAG)

- DAGs consist of **variables (nodes)** and **arrows (edges)**

- Nodes can represent treatments, exposures, outcomes, or other characteristics

- Arrows represent **known or hypothesised causal relationships** between variables

## Direct Acyclic Graph (DAG)

DAGs depict an investigator’s hypotheses or assumptions about:

- how variables are causally related

- how an exposure/intervention affects an outcome

- which variables should be controlled for (in other words, \*\*DAGs can inform statistical analysis\*\*)

## Types of variables

When deciding which variables to include in a regression model, it is important to distinguish between:

- **Predictors**

- **Mediators**

- **Confounders**

- **Colliders**

These variables have very different implications for statistical adjustment.

## Predictors

- A predictor ($X$) is a variable included in the model to explain the outcome ($Y$)

## Predictors in a DAG

```{r}
coord_dag <- list(
  x = c(X1 = 0, X2 = 0, Y = 2),
  y = c(X1 = 0, X2 = 1, Y = 0)
)

our_dag <- ggdag::dagify(Y ~ X1,
                         Y ~ X2,
                         coords = coord_dag,
                         labels = c("X1" = "Sleep time",
                                    "X2" = "Study time",
                                    "Y" = "Grade")
                         )

ggdag(our_dag,
      use_labels = "label") + theme_void()
```

​This represents: $X1​$ → $Y$ and $X2$ ​→ $Y$

## Simulating predictors

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(sleep_time = rnorm(100, 10, 2),
                   study_time = rnorm(100, 3, 3)) |>
  mutate(grades = 7 + 0.4 * sleep_time + 0.7 * study_time + rnorm(100, 0, 3))

# Display first 4 rows, round to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(1:3), digits = 2)
```

## Regression model

```{r}
#| echo: TRUE

# Run regression analysis and display output, rounded to 2 decimals
lm(grades ~ sleep_time + study_time, data = data) |>
  tidy() |>
  select(term, estimate, p.value) |>
  flextable() |>
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "blue", part = "body") |> 
  color(i = 3, color = "green", part = "body")
```

## Interpreting the predictors

> **Sleep time:** Holding study time constant, a one-unit increase in sleep time is associated with a **0.4-unit increase in grades**.

> **Study time:** Holding sleep time constant, a one-unit increase in study time is associated with a **0.7-unit increase in grades**.

## Mediator

A **mediator** is a variable that:

- is influenced by the exposure ($X$)

- influences the outcome ($Y$)

- lies on the causal pathway between X and Y

- controlling for it removes the **mediated effect** from the estimated effect of $X$

## Mediator in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, M = 1, Y = 2),
  y = c(X = 0, M = 1, Y = 0)
)


our_dag <- dagify(M ~ X,
                  Y ~ M,
                  Y ~ X,
                  coords = coord_dag,
                  labels = c("M" = "Fast food", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Mortality")
                  )

ggdag(our_dag, 
      use_labels = "label") + theme_void()
```

## Direct, Indirect and Total Effects

**Direct effect**

- The effect of $X$ on $Y$ not through $M$ : $X$ → Y

**Indirect effect**

- The effect of $X$ on $Y$ through $M$: $X$ → $M$ → $Y$

**Total effect**

The overall effect of $X$ on $Y$, combining the direct and indirect effects:

- Total effect = Direct effect + Indirect effect

## Simulating a mediator

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  exposure = rnorm(100)
) |>
  mutate(
    mediator = 0.6 * exposure + rnorm(100, 0, 1),
    outcome = 0.5 * exposure + 0.7 * mediator + rnorm(100, 0, 1)
  )

# Display the first 4 observations, rounded to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## Without controlling for the mediator

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data = data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2)
```

This estimates the total effect of the exposure. This is the effect of interest when estimating the overall effect of an exposure or intervention.

## Controlling for the mediator

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + mediator, data = data) |>
  tidy() |>
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

This estimates the direct effect of the exposure (i.e., the effect of $X$ on $Y$ **not through M**).

Conditioning on the mediator blocks the mediated (indirect) effect and can therefore bias estimates of the total effect of $X$ on $Y$

## Confounder

A **confounder** is a variable that:

- is associated with the exposure ($X$)
- is associated with the outcome ($Y$)
- is not a consequence of the exposure
- can lead to biased estimates if not controlled for

## Confounder in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(X ~ C, 
                  Y ~ X, 
                  Y ~ C, 
                  coords = coord_dag,
                  labels = c("C" = "SEP*", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Mortality")
                  )

ggdag(our_dag,
      use_labels = "label") + theme_void()
```

\*SEP = socio-economic position

## Confounding in practice

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  covariate = sample(0:1, 100, replace = TRUE)
) |>
  dplyr::mutate(
    exposure = runif(100, 0, 1) + 0.3 * covariate,
    outcome = 2.0 + 0.5 * exposure + 0.25 * covariate
  )

# Display first 4 rows, rounded to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2)
```

## Without adjusting for the confounder

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Without adjusting for the confounder, the estimated effect is a biased estimate of the true effect size

## Adjusting for the confounder

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + covariate, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "blue", part = "body")
```

Adjusting for the confounder results in an unbiased estimate of effect of the exposure

## Collider

A **collider** is a variable that:

- is caused by the exposure ($X$)

- is caused by the outcome ($Y$)

- is therefore a common consequence of both the exposure and the outcome

- Controlling for a collider can induce a spurious association between $X$ and $Y$

Collider bias can arise through:

- Study design: sampling, stratification or matching

- Statistical analysis: adjusting for the collider in a regression model

## An example: The Obesity Paradox

- In the general population, higher BMI (above \~30 kg/m²) is generally associated with higher mortality risk

- Previous research investigated whether obesity increase mortality among people with CVD

- However, CVD can act as a collider because both obesity and other causes of mortality can contribute to CVD

- **Conditioning on CVD** (e.g., by stratifying on disease status or by adjusting on multiple regression) can create a spurious association between obesity and mortality

- This can produce the **obesity paradox**: among people with CVD, obesity may appear to be associated with **lower mortality**, despite being associated with higher mortality in the general population

## Collider in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(C ~ X, 
                  Y ~ X, 
                  C ~ Y, 
                  coords = coord_dag,
                  labels = c("C" = "CVD",
                             "X" = "Obesity",
                             "Y" = "Mortality")
)

ggdag(our_dag,
      use_labels = "label") + 
  theme_void()
```

\*CVD = Cardiovascular disease

## An example: The Obesity Paradox

| **Analysis** | **RD (95% CI)** |
|:--------------------------------|:--------------------------------------|
| Crude | 0.05 (0.03, 0.07) |
| Adjusted — total population | 0.03 (0.02, 0.05) |
| [Adjusted — CVD]{style="color:red"} | [−0.12 (−0.20, −0.04)]{style="color:red"} |
| Adjusted — No CVD | 0.03 (0.01, 0.05) |

: Comparison of the estimated risk difference for the effect of obesity on mortality. Results obtained from @obesity_paradox2015

## Simulating a collider

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  exposure = rnorm(1000),
  outcome = rnorm(1000)
) |>
  mutate(
    collider = exposure + outcome + rnorm(1000, 0, 0.5)
  )

# Display first 4 rows, rounded to 2 DECIMALS
data |> 
  head(4) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## No conditioning on a collider

- Exposure and outcome were simulated **independently**.

- Therefore, there is **no true association** between them.

```{r}
#| echo: TRUE
#| out-width: 100%

ggplot(data, aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## No adjustement for collider

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

*No* association between exposure and outcome.

## Conditioning on a collider through sampling

```{r}
#| echo: TRUE
data |> 
  filter(collider > 0.2) |>  # Select observations based on the collider
  ggplot(aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## Adjusting for a collider

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + collider, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Controlling for the collider creates a spurious association between exposure and outcome

## Confirmatory research

- Aims to answer a **well-defined, pre-specified question**:\
  *“Does heavy alcohol consumption increase mortality?”*

- Focuses on estimating or testing a **specific relationship** between variables

- Predictors are selected based on **theory, prior evidence, or causal assumptions**

- When estimating a causal effect, the study design and analysis should account for **confounding** (e.g., through randomization or appropriate adjustment)

- An observational study can be **confirmatory** if the research question, design, and analysis are **pre-specified and appropriately justified**

## Exploratory research

## Exploratory research

- Does not have a **specific pre-specified hypothesis**

- 

  - Research questions often use wording such as:\
    *“What factors are associated with mortality?”*

- Examines multiple variables to identify **potential associations**

- Findings are **hypothesis-generating** and should be confirmed in future research

- Observational studies can be affected by **unmeasured confounding, unadjusted confounding, and selection bias**

## Prediction

- Asks: “How accurately can we predict $Y$?”

- Aims to build the best-performing model using available predictors

- Uses observed data to make predictions about an outcome

- Focuses on prediction accuracy, rather than estimating causal effects

- The best model is the one that generalizes well to new, unseen data

## Recipe for a good confirmatory study

::: callout-tip
- **Ask a clear causal question.** Use unambiguous causal language, such as: “Does reducing salt intake prevent strokes?”

- **Prespecify the adjustment strategy.** Base your strategy on existing evidence, substantive knowledge, and explicit assumptions. A **directed acyclic graph (DAG)** can help communicate which variables should—and should not—be adjusted for.

- **Adjust for confounders.** Include the variables needed to block relevant backdoor paths between the exposure and outcome in the regression model.

- **Avoid inappropriate adjustment.** Do not adjust for **mediators** or **colliders**, as this can bias the estimated causal effect.
:::

## References and further reading

Further reading includes @tutorial_dag2022 and @lewer_2025
