---
title: "Multiple Regression I"
subtitle: "Variable Selection"
author: "Cristian Mesquida"
institute: "Eindhoven University of Technology"
date: today
format: revealjs
slide-number: true
bibliography: references.bib
nocite: |
 @*
editor: visual
css: styles.css
---

```{r packages}
#| include: false

library(dplyr)
library(ggplot2)
library(broom)
library(ggdag)
library(flextable)
```

```{r setup}
#| include: false

set_flextable_defaults(
  font.size = 20
)
```

## Abstract

When conducting a multiple regression, researchers face an important question: **What variables should I include in the model?** This lecture introduces variable selection in multiple regression, with a focus on choosing predictors based on the research question. It first introduces different types of research questions, including causal, associational, and prediction questions. It then introduces Directed Acyclic Graphs (DAGs) and shows how they can help identify potential confounders and avoid inappropriate adjustment for mediators and colliders.

## Lecture's objectives

By the end of this lecture, you should be able to:

- Identify the type of research question (causal, associational or prediction)

- Identify the type of variables (predictors, mediators, confounders and colliders)

- Be familiar with Directed Acyclical Graphs (DAGs) and how they can help identify potential confounders and avoid inappropriate adjustment for mediators and colliders

## Simple regression

The simplest form of regression analysis can be expressed as:

$$Y_i = \beta_0 + \beta_1X_{1i} + \epsilon_i $$

Where $Y_i$ = observed outcome, $\beta_0$ = intercept, $\beta_1$ = regression coefficient, $X_{1i}$ = predictor and $\epsilon_i$ = error term for observation $i$

## Simple regression

- It examines how a single independent variable predicts a single dependent variable.

- For example, we can use regression to answer the research question: **Does exercise reduces blood pressure?**

$$\text{blood_pressure}_i = \beta_0 + \beta_1  \text{exercise} + \epsilon_i $$

Where **Exercise** is an categorical intervention variable (e.g., 0 = no exercise and 1 = exercise)

## Multiple regression

When the model includes more than one predictor, it is called **multiple** or **multivariable** regression. It can be expressed as:

$$Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + ... +  \epsilon_i$$

Where $\beta_1, \beta_2, \dots$ = regression coefficients and $X_{1i}, X_{2i}, \dots$ = predictors and and $\epsilon_i$ = error term for observation $i$

## Multiple regression

- It examines how a dependent variable is related to multiple independent variables, while **adjusting or controlling for** the other predictors in the model.

- For example, we know that **cholesterol** and **age** may also influence **blood pressure**, so we include them in the model:

$$\text{blood_pressure}_i = \beta_0 + 
\beta_1  \text{exercise} + 
\beta_2  \text{cholesterol} + 
\beta_3 \text{age} + 
\epsilon $$

Where **exercise** is categorical (e.g., coded as 0/1) and **cholesterol** and **age** are a continuous variable

## Example

```{r}
nhefs <- readr::read_csv(
  "https://vincentarelbundock.github.io/Rdatasets/csv/causaldata/nhefs.csv",
  show_col_types = FALSE
)
```

- Now, let's suppose we are researchers working for the Ministry of Health, and we want to investigate the effect of an intervention on a health outcome.

- We will use the `nhefs` dataset as example. It contains data from [The National Health and Nutrition Examination Survey Data I Epidemiologic Follow-up Study (NHEFS)](https://vincentarelbundock.github.io/Rdatasets/articles/data.html).

- The original dataset has `r dim(nhefs)[[1]]` rows and `r dim(nhefs)[[2]]` variables.

- Such a rich dataset allows us to conduct many analyses—but which variables should we include in our model?

```{r}
#| eval: FALSE
nhefs |> 
  select(3:13) |> 
  slice_head(n = 5) |> 
  flextable()
```

## What variables to include in the model?

::: {.fragment .fade-in-then-out}
**Not** based on whether *p* \< 0.05!

Selecting variables based solely on statistical significance is **a mindless approach** and should be avoided

It can increase the number of false positives (Type I error rate)
:::

::: {.fragment .fade-up}
The answer is that it depends on your **research question**
:::

# Research questions

## Causal questions

- Aims to answer a **well-defined, pre-specified question**: *“Does heavy alcohol consumption increase mortality?”*

- Focus on estimating the causal effect of an an intervention/exposure on an outcome

- Variables should be selected based on causal assumptions, theory, and prior evidence

- Confounding needs to be addressed through study design and/or appropriate adjustment

## Causal questions

- Causal questions can be studied using randomised controlled trials (RCTs) or observational studies

- RCTs are generally preferred for establishing causal effects but they can be costly or unethical (e.g., randomly assigning people to smoke to study its effect on lung cancer)

- Observational studies can also support causal inference, but confounding and other sources of bias are important limitations

- Regression analysis is commonly used in observational studies because it can adjust for measured confounders by including them as covariates in the model

## Exploratory /associational questions

- Does not have a **specific pre-specified hypothesis**

- Research questions often use wording such as:*“What factors are associated with mortality?”*

- Examines multiple variables to identify **potential associations**

- Findings are **hypothesis-generating** and should be confirmed in future research

- This type of study can suffer from confounding and other source of bias, so causal claims should be avoided

## Prediction questions

- Asks: “How accurately can we predict risk of mortality in patients suffering from metabolic syndrome?”

- Aims to build the best-performing model using available predictors

- Uses observed data to make predictions about an outcome

- Focuses on prediction accuracy, rather than estimating causal effects

- The best model is the one that generalizes well to new, unseen data

## Causal questions and DAGs

```{r}
dag <- dagify(
  bp ~ obesity + sep + exercise,
  obesity ~ exercise,
  hospialization ~ exercise + bp,
  exercise ~ sep,
  
  labels = c(
    "bp" = "Blood\n Pressure",
    "obesity" = "Obesity",
    "sep" = "SEP",
    "hospialization" = "Hospialization",
    "exercise" = "Physical\n Activity"
  )
)

ggdag(
  dag,
  text = FALSE,
  use_labels = "label"
) +
  theme_void()
```

## Directed Acyclic Graph (DAG)

- DAGs consist of **variables (nodes)** and **arrows (edges)**

- Nodes can represent interventions, exposures, outcomes, or other characteristics

- Arrows represent **known or hypothesised causal relationships** between variables

## Directed Acyclic Graph (DAG)

DAGs depict an investigator’s hypotheses or assumptions about:

- how variables are causally related

- how an intervention/exposure affects an outcome

- which variables should be adjusted for (in other words, **DAGs can inform statistical analysis**)

## Types of variables

When deciding which variables to include in a regression model, it is important to distinguish between:

- **Predictors**

- **Mediators**

- **Confounders**

- **Colliders**

These variables have very different implications for statistical adjustment.

## Predictors

- A predictor ($X$) is a variable included in the model to explain the outcome ($Y$)

## Predictors in a DAG

```{r}
coord_dag <- list(
  x = c(X1 = 0, X2 = 0, Y = 2),
  y = c(X1 = 0, X2 = 1, Y = 0)
)

our_dag <- ggdag::dagify(Y ~ X1,
                         Y ~ X2,
                         coords = coord_dag,
                         labels = c("X1" = "Physical\n Activity",
                                    "X2" = "Food\n Intake",
                                    "Y" = "Blood\n Pressure")
                         )

ggdag(our_dag,
      use_labels = "label") + theme_void()
```

## Simulating predictors

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(food_intake = rnorm(200, 10, 2),
                   exercise = rnorm(200, 3, 3)) |>
  mutate(blood_pressure = 7 + 0.4 * food_intake + 0.7 * exercise + rnorm(200, 0, 3))

# Display first 4 rows, round to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(1:3), digits = 2)
```

## Regression model

```{r}
#| echo: TRUE

# Run regression analysis and display output, rounded to 2 decimals
lm(blood_pressure ~ food_intake + exercise, data = data) |>
  tidy() |>
  select(term, estimate, p.value) |>
  flextable() |>
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "blue", part = "body") |> 
  color(i = 3, color = "green", part = "body")
```

## Interpreting the predictors

> **Food intake:** Holding `exercise` constant, a one-unit increase in diet is associated with a **0.4-unit increase in blood pressure**.

> **Exercise:** Holding `food_intake` constant, a one-unit increase in `exercise` is associated with a **0.7-unit increase in blood pressure**.

## Mediator

A **mediator** is a variable that:

- is influenced by the exposure ($X$)

- influences the outcome ($Y$)

- lies on the causal pathway between X and Y

- adjusting for it removes the **mediated effect** from the estimated effect of $X$

## Mediator in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, M = 1, Y = 2),
  y = c(X = 0, M = 1, Y = 0)
)


our_dag <- dagify(M ~ X,
                  Y ~ M,
                  Y ~ X,
                  coords = coord_dag,
                  labels = c("M" = "Fast food", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Mortality")
                  )

ggdag(our_dag, 
      use_labels = "label") + theme_void()
```

## Direct, Indirect and Total Effects

**Direct effect**

- The effect of $X$ on $Y$ not through $M$ : $X$ → Y

**Indirect effect**

- The effect of $X$ on $Y$ through $M$: $X$ → $M$ → $Y$

**Total effect**

The overall effect of $X$ on $Y$, combining the direct and indirect effects:

- Total effect = Direct effect + Indirect effect

## Simulating a mediator

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  exposure = rnorm(200)
) |>
  mutate(
    mediator = 0.6 * exposure + rnorm(200, 0, 1),
    outcome = 0.5 * exposure + 0.7 * mediator + rnorm(200, 0, 1)
  )

# Display the first 4 observations, rounded to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## Without adjustment regression of the mediator

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data = data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "green", part = "body")
```

This estimates the total effect of the exposure. This is the effect of interest when estimating the overall effect of an exposure or intervention.

## With adjustment regression of the mediator

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + mediator, data = data) |>
  tidy() |>
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

This estimates the direct effect of the exposure (i.e., the effect of $X$ on $Y$ **not through M**).

Adjusting for the mediator blocks the pathway $X$ → $M$ → $Y$, so the model no longer estimates the total effect of $X$ on $Y$.

## Confounder

A **confounder** is a variable that:

- is associated with the exposure ($X$)
- is associated with the outcome ($Y$)
- is not a consequence of the exposure
- can lead to biased estimates if not adjusted for

## Confounder in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(X ~ C, 
                  Y ~ X, 
                  Y ~ C, 
                  coords = coord_dag,
                  labels = c("C" = "SEP*", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Mortality")
                  )

ggdag(our_dag,
      use_labels = "label") + theme_void()
```

\*SEP = socio-economic position

## Confounding in practice

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  covariate = sample(0:1, 100, replace = TRUE)
) |>
  dplyr::mutate(
    exposure = runif(100, 0, 1) + 0.3 * covariate,
    outcome = 2.0 + 0.5 * exposure + 0.25 * covariate
  )

# Display first 4 rows, rounded to 2 decimals
head(data, 4) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2)
```

## Without adjustment regression of the confounder

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Without adjusting for the confounder, the estimated effect is a biased estimate of the true effect size

## With regression adjustment of the confounder

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + covariate, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "blue", part = "body")
```

Adjusting for the confounder results in an unbiased estimate of effect of the exposure

## Collider

A **collider** is a variable that:

- is caused by the exposure ($X$)

- is caused by the outcome ($Y$)

- is therefore a common consequence of both the exposure and the outcome

- Adjusting for a collider can induce a spurious association between $X$ and $Y$

## 

Collider bias can arise through:

- Study design: sampling, stratification or matching

- Statistical analysis: adjusting for the collider in a regression model

## An example: The Obesity Paradox

- In the general population, the higher the body mass index higher the mortality risk

- Previous research investigated whether obesity increase mortality among people with cardiovascular disease (CVD)

- However, CVD can act as a collider because both obesity and other causes of mortality can contribute to CVD

- **Conditioning on CVD** (e.g., by stratifying on disease status or by adjusting on multiple regression) can create the **obesity paradox**

- **Obesity paradox:** among people with CVD, obesity may appear to be associated with lower mortality, despite being associated with higher mortality in the general population

## Collider in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(C ~ X, 
                  Y ~ X, 
                  C ~ Y, 
                  coords = coord_dag,
                  labels = c("C" = "CVD",
                             "X" = "Obesity",
                             "Y" = "Mortality")
)

ggdag(our_dag,
      use_labels = "label") + 
  theme_void()
```

\*CVD = Cardiovascular disease

## The Obesity Paradox

| **Analysis** | **RD (95% CI)** |
|:--------------------------------|:--------------------------------------|
| Crude | 0.05 (0.03, 0.07) |
| Adjusted — total population | 0.03 (0.02, 0.05) |
| [Adjusted — CVD]{style="color:red"} | [−0.12 (−0.20, −0.04)]{style="color:red"} |
| Adjusted — No CVD | 0.03 (0.01, 0.05) |

: Comparison of the estimated risk difference for the effect of obesity on mortality. Results obtained from @obesity_paradox2015

## Simulating a collider

```{r}
#| echo: TRUE

set.seed(1234)

# Simulate data
data <- data.frame(
  exposure = rnorm(1000),
  outcome = rnorm(1000)
) |>
  mutate(
    collider = exposure + outcome + rnorm(1000, 0, 0.5)
  )

# Display first 4 rows, rounded to 2 DECIMALS
data |> 
  head(4) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## No conditioning on a collider

- Exposure and outcome were simulated **independently**.

- Therefore, there is **no true association** between them.

```{r}
#| echo: TRUE
#| out-width: 100%

ggplot(data, aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## Without regression adjustment of the collider

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

*No* association between exposure and outcome.

## Conditioning on a collider through sampling

```{r}
#| echo: TRUE
data |> 
  filter(collider > 0.2) |>  # Select observations based on the collider
  ggplot(aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## With regression adjustment of the collider

```{r}
#| echo: TRUE

# Run model and display estimates and p-values, rounded to 2 decimals
lm(outcome ~ exposure + collider, data) |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Adjusting for the collider creates a spurious association between exposure and outcome

## Recipe for a good confirmatory study

::: callout-tip
- **Ask a clear causal question.** Use unambiguous causal language, such as: “Does reducing salt intake prevent strokes?”

- **Prespecify the adjustment strategy.** Base your strategy on existing evidence, substantive knowledge, and explicit assumptions. A **directed acyclic graph (DAG)** can help communicate which variables should—and should not—be adjusted for.

- **Regression adjustment of confounders.** Include the variables needed to block relevant backdoor paths between the exposure and outcome in the regression model.

- **Avoid inappropriate adjustment.** Do not adjust for **mediators** or **colliders**, as this can bias the estimated causal effect.
:::

## References and further reading

Further reading includes @tutorial_dag2022 and @lewer_2025
