---
title: "Multiple Regression I: Variable Selection"
subtitle: "Behavioral Research Methods 2"
institute: "Eindhoven University of Technology"
author: "Cristian Mesquida"
date: today
format: revealjs
slide-number: true
bibliography: references.bib
nocite: |
 @*
editor: visual
css: styles.css
---

```{r packages}
#| include: false

library(dplyr)
library(ggplot2)
library(broom)
library(ggdag)
library(flextable)
```

```{r setup}
#| include: false

set_flextable_defaults(
  font.size = 20
)
```

## Lecture's objectives

By the end of this lecture, you should be able to:

- Identify the type of research question (causal, associational or prediction)

- Identify the type of variables (predictors, mediators, confounders and colliders)

- Be familiar with Directed Acyclical Graphs (DAGs) and how they can help identify potential confounders and avoid inappropriate adjustment for mediators and colliders

## First

```{r}
nhefs <- readr::read_csv(
  "https://vincentarelbundock.github.io/Rdatasets/csv/causaldata/nhefs.csv",
  show_col_types = FALSE
)
```

- Suppose we are researchers working for the Ministry of Health

- We want to investigate the predictors of blood pressure (`sbp`)

- We use the [`nhefs`](https://vincentarelbundock.github.io/Rdatasets/articles/data.html) dataset, which contains [survey data](https://wwwn.cdc.gov/Nchs/Nhanes/nhefs/default.aspx) from US population

- Which variables predict systolic blood pressure (`sbp`)?

::: scrollable
```{r}
nhefs |> 
  slice_head(n = 5) |> 
  flextable() |> 
  color(j = 8, color = "blue", part = "all")
```
:::

## What variables to include in the model?

::: {.fragment .fade-in-then-out}
**Not** based on whether *p* \< 0.05!

Selecting variables based solely on statistical significance is **a mindless approach** and should be avoided

It can increase the number of false positives (Type I error rate)
:::

::: {.fragment .fade-up}
The answer is that it depends on your **research question**
:::

# Types of research questions

## Causal questions

- Aims to answer a **well-defined, pre-specified question**: *“Does heavy sodium intake increase blood pressure?”*

- Focus on estimating the causal effect of an an intervention/exposure on an outcome

- Variables should be selected based on causal assumptions, theory, and prior evidence

## Causal questions

- Causal questions can be studied using **randomised controlled trials** (RCTs) or **observational studies**

- RCTs are generally preferred for establishing causal effects but they can be costly or unethical (e.g., randomly assigning people to smoke to study its effect on lung cancer)

- Observational studies can also support causal inference, but confounding needs to be addressed

- Regression analysis can be used to adjust for measured confounders by including them as covariates in the model

## A side note on **observational studies**

- Observational studies are based on data collected without any intervention or manipulation by the researcher

- Observational studies involve collecting data from:

  - Medical records, surveys, registries, etc.

  <!-- -->

  - Observing participants in their natural environment

- Observational studies can be:

  - **Prospective** (data collected going forward in time)

  - **Retrospective** (data collected going backwards in time)

## Exploratory / associational questions

- Does not have a **specific pre-specified hypothesis**

- Research questions often use wording such as:*“What factors are associated with mortality?”*

- Examines multiple variables to identify **potential associations**

- Based on observational studies so there is risk of confounding

## Prediction questions

- Asks: “*How accurately can we predict blood pressure in the sedentary population?*”

- Uses observed data to make predictions about an outcome

- Focuses on prediction accuracy, rather than estimating causal effects

- The best model is the one that generalizes well to new, unseen data

## Directed Acyclic Graph (DAG)

- DAGs consist of **variables (nodes)** and **arrows (edges)**

- Nodes can represent interventions, exposures, outcomes, or other characteristics

- Arrows represent **known or hypothesised causal relationships** between variables

## Directed Acyclic Graph (DAG)

DAGs depict an investigator’s hypotheses or assumptions about:

- how variables are causally related

- how an intervention/exposure affects an outcome

- which variables should be adjusted for (in other words, **DAGs can inform statistical analysis**)

## Types of variables

When deciding which variables to include in a regression model, it is important to distinguish between:

- **Predictors**

- **Mediators**

- **Confounders**

- **Colliders**

These variables have very different implications for statistical adjustment.

## Predictors

- A predictor ($X$) is a variable included in the model to explain the outcome ($Y$)

- Predictors can be:

  - Exposures/interventions (e.g., diet, exercise, medication, etc.)
  - Covariates: other characteristics that may influence the outcome (e.g., age, sex, socio-economic position, etc.)

## Predictors in a DAG

```{r}
coord_dag <- list(
  x = c(X1 = 0, X2 = 0, Y = 2),
  y = c(X1 = 0, X2 = 1, Y = 0)
)

our_dag <- dagify(
  Y ~ X1,
  Y ~ X2,
  coords = coord_dag,
  labels = c("X1" = "Physical\n Activity",
             "X2" = "Sodium\n Intake",
             "Y" = "Blood\n Pressure")
  )

ggdag(our_dag, use_labels = "label") + theme_void()
```

## Simulated example of predictors

```{r}
#| echo: FALSE

set.seed(1234)

# Simulate data
data <- data.frame(sodium_intake = rnorm(200, 10, 2),
                   exercise = rnorm(200, 3, 3)) |>
  mutate(blood_pressure = 7 + 1 * sodium_intake + -1.7 * exercise + rnorm(200, 0, 3))

# Display first 8 rows, round to 2 decimals
head(data, 8) |> 
  flextable() |> 
  colformat_double(j = c(1:3), digits = 2)
```

## Regression model

```{r}
#| echo: TRUE
res <- lm(blood_pressure ~ exercise + sodium_intake, data = data)
```

```{r}
#| echo: FALSE
res |>
  tidy() |>
  select(term, estimate, p.value) |>
  flextable() |>
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "blue", part = "body") |> 
  color(i = 3, color = "green", part = "body")
```

## Interpreting the predictors

> **Sodium intake:** Holding `exercise` constant, a one-unit increase in sodium is associated with a **`r round(res$coefficients[2], 2)`-unit decrease in blood pressure**\*

> **Exercise:** Holding `sodium_intake` constant, a one-unit increase in `exercise` is associated with a **`r round(res$coefficients[3], 2)`-unit increase in blood pressure**

## Mediator

A **mediator** is a variable that:

- is influenced by the exposure ($X$)

- influences the outcome ($Y$)

- lies on the causal pathway between $X$ and $Y$

- adjusting for it removes the **mediated effect** from the estimated effect of $X$

## Mediator in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, M = 1, Y = 2),
  y = c(X = 0, M = 1, Y = 0)
)


our_dag <- dagify(M ~ X,
                  Y ~ M,
                  Y ~ X,
                  coords = coord_dag,
                  labels = c("M" = "Sodium\n intake ", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Blood\n pressure")
                  )

ggdag(our_dag, 
      use_labels = "label") + theme_void()
```

## Direct, Indirect and Total Effects

**Direct effect**

- The effect of $X$ on $Y$ not through $M$ : $X$ → Y

**Indirect effect**

- The effect of $X$ on $Y$ through $M$: $X$ → $M$ → $Y$

**Total effect**

The overall effect of $X$ on $Y$, combining the direct and indirect effects:

- Total effect = Direct effect + Indirect effect

## Simulated example of a mediator

```{r}
#| echo: FALSE

set.seed(1234)

# Simulate data
data <- data.frame(
  exposure = rnorm(200)
) |>
  mutate(
    mediator = 0.6 * exposure + rnorm(200, 0, 1),
    outcome = 0.5 * exposure + 0.7 * mediator + rnorm(200, 0, 1)
  )

# Display the first 8 observations, rounded to 2 decimals
head(data, 8) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## Without adjusting for the mediator

```{r}
#| echo: TRUE
res <- lm(outcome ~ exposure, data = data) 
```

```{r}
#| echo: FALSE
res |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "green", part = "body")
```

This model estimates the total effect of the [exposure]{style="color:green;"}.

The total effect is the effect of interest when estimating the overall effect of an exposure or intervention.

## Adjusting for the mediator

```{r}
#| echo: TRUE

res <- lm(outcome ~ exposure + mediator, data = data)
```

```{r}
#| echo: FALSE
res |>
  tidy() |>
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

[exposure]{style="color:red;"} corresponds to the direct effect (i.e., the effect of $X$ on $Y$ **not through M**).

Adjusting for the mediator blocks the pathway $X$ → $M$ → $Y$, so the model no longer estimates the total effect of $X$ on $Y$

## Confounder

A **confounder** is a variable that:

- is associated with the exposure ($X$)
- is associated with the outcome ($Y$)
- is not a consequence of the exposure
- can lead to biased estimates if not adjusted for

## Confounder in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(X ~ C, 
                  Y ~ X, 
                  Y ~ C, 
                  coords = coord_dag,
                  labels = c("C" = "SEP*", 
                             "X" = "Heavy\n alcohol",
                             "Y" = "Blood\n pressure")
                  )

ggdag(our_dag,
      use_labels = "label") + theme_void()
```

\*SEP = socio-economic position

## Confounding in practice

```{r}
#| echo: FALSE

set.seed(1234)

# Simulate data
data <- data.frame(
  covariate = sample(0:1, 100, replace = TRUE)
) |>
    mutate(
    exposure = runif(100, 0, 1) + 0.3 * covariate,
    outcome = 2.0 + 0.5 * exposure + 0.25 * covariate
  )

# Display first 8 rows, rounded to 2 decimals
head(data, 8) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2)
```

## Without adjusting for the confounder

```{r}
#| echo: TRUE
res <- lm(outcome ~ exposure, data) 
```

```{r}
#| echo: FALSE
res |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Without adjusting for the confounder, the estimated effect of [exposure]{style="color:red;"} is biased

## Adjusting for the confounder

```{r}
#| echo: FALSE
res <- lm(outcome ~ exposure + covariate, data)
```

```{r}
res |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "green", part = "body")
```

Adjusting for the confounder results in an unbiased estimate of effect of the [exposure]{style="color:green;"}

## Collider

A **collider** is a variable that:

- is caused by the exposure ($X$)

- is caused by the outcome ($Y$)

- is therefore a common consequence of both the exposure and the outcome

- Collider bias can arise through:

  - Study design: sampling, stratification or matching
  - Statistical analysis: adjusting for the collider in a regression model

## Collider in a DAG

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1, Y = 0)
)
our_dag <- dagify(C ~ X, 
                  Y ~ X, 
                  C ~ Y, 
                  coords = coord_dag,
                  labels = c("C" = "CVD",
                             "X" = "Obesity",
                             "Y" = "Mortality")
)

ggdag(our_dag,
      use_labels = "label") + 
  theme_void()
```

\*CVD = Cardiovascular disease

## Simulated example of collider bias

```{r}
#| echo: FALSE

set.seed(1234)

# Simulate data
data <- data.frame(exposure = rnorm(1000),
                   outcome = rnorm(1000)) |>
  mutate(collider = exposure + outcome + rnorm(1000, 0, 0.5))

# Display first 8 rows, rounded to 2 decimals
data |> 
  head(8) |> 
  flextable() |> 
  colformat_double(j = c(1, 2, 3), digits = 2)
```

## No conditioning on a collider

- Exposure and outcome were simulated **independently**

- Therefore, there is **no true association** between them

```{r}
#| echo: FALSE
#| out-width: 100%

ggplot(data, aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## Without adjusting for the collider

```{r}
#| echo: TRUE
res <- lm(outcome ~ exposure, data)
```

```{r}
#| echo: FALSE
res |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "green", part = "body")
```

[No causal association]{style="color:green;"} between [exposure]{style="color:green;"} and outcome

## Conditioning on a collider through sampling

```{r}
#| echo: FALSE
data |> 
  filter(collider > 0.2) |>  # Select observations based on the collider
  ggplot(aes(x = exposure, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

## Adjusting for the collider

```{r}
#| echo: TRUE
res <- lm(outcome ~ exposure + collider, data)
```

```{r}
#| echo: FALSE
res |> 
  tidy() |> 
  select(term, estimate, p.value) |> 
  flextable() |> 
  colformat_double(j = c(2, 3), digits = 2) |> 
  color(i = 2, color = "red", part = "body")
```

Adjusting for the collider creates [a spurious association]{style="color:red;"} between [exposure]{style="color:red;"} and outcome

## An example: The Obesity Paradox

- The higher the body mass index, the higher the mortality risk

- Cardiovascular disease (CVD) can act as a collider

- **Conditioning on CVD** (e.g., by stratifying on disease status or by adjusting on multiple regression) create the **obesity paradox**

- **Obesity paradox:** among people with CVD, obesity appear to be associated with lower mortality

## The Obesity Paradox

| **Analysis** | **Risk Difference (95% CI)** |
|:--------------------------------|:--------------------------------------|
| [Crude (mortality \~ obesity + confounders)]{style="color:green"} | [0.05 (0.03, 0.07)]{style="color:green"} |
| Adjusted — total population | 0.03 (0.02, 0.05) |
| [Adjusted — CVD (mortality \~ obesity + CVD + confounders)]{style="color:red"} | [−0.12 (−0.20, −0.04)]{style="color:red"} |
| Adjusted — No CVD | 0.03 (0.01, 0.05) |

: Comparison of the estimated risk difference for the effect of obesity on mortality. Results obtained from @obesity_paradox2015

## Another example of collider bias

```{r}
coord_dag <- list(
  x = c(X = 0, C = 1, Y = 2),
  y = c(X = 0, C = 1.5, Y = 0)
)

our_dag <- dagify(
  C ~ X + Y,
  Y ~ X,
  coords = coord_dag,
  labels = c(
    C = "University enrollment",
    X = "Sport ability",
    Y = "Academic ability"
  )
)

ggdag(our_dag) +
  annotate("text", x = 1, y = 1.25, label = "University enrollment") +
  annotate("text", x = 0, y = -0.25, label = "Sport ability") +
  annotate("text", x = 2, y = -0.25, label = "Academic ability") +
  theme_void()
```

## Conditioning on a collider through sampling

```{r}
data <- data.frame(
  exposure = rnorm(1000, mean = 5, sd = 1.4),
  outcome  = rnorm(1000, mean = 5, sd = 1.4)
) |>
  mutate(
    collider = 3 * exposure + outcome + rnorm(1000, 0, 0.2),
    group = ifelse(
      collider > 20,
      "Selected observations",
      "Non-selected observations"
    ),
    group = factor(
      group,
      levels = c("Non-selected observations", "Selected observations")
    )
  )

ggplot(data, aes(exposure, outcome)) +
  geom_point(alpha = 0.2) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 1
  ) +
  geom_point(
    data = subset(data, group == "Selected observations"),
    aes(colour = group),
    alpha = 0.5
  ) +
  geom_smooth(
    data = subset(data, group == "Selected observations"),
    method = "lm",
    se = FALSE,
    aes(colour = group),
    linewidth = 1.2
  ) +
  ylab("Sport ability") +
  xlab("Academic ability") +
  theme_classic()
```

Conditioning on the collider (i.e., university enrollment) creates a spurious association between sports ability and academic ability

## Recipe for an appropriate regression model

- **Start with the research question** — what effect or association are you trying to estimate?

- **Adjust for relevant measured confounders** that are common causes of the exposure and outcome

- **For the total effect of an intervention/exposure, do not adjust for mediators** that lie on the causal pathway

- **Do not condition on colliders**, either through sampling/selection or regression adjustment

## Further reading and references

Further reading includes @tutorial_dag2022 and @lewer_2025
